On the uniqueness and stability of dictionaries for sparse representation of noisy signals

### 2017-08-21 manuscripts/PNAS/review/point_by_point_response.txt

References for Cover Letter and Point-by-point Response:

(IHS10) Isely, Hillar, Sommer, Advances in neural information processing systems (NIPS), 2010.
(M10) Morris I (2010), Advances in Mathematics 225(6):3425–3445.
(GS12) Ganguli, Sompolinksi, Annual review of neuroscience, 2012.
(ACD13) Arias-Castro, Candes, Davenport, IEEE Trans. Inf. Theory, 2013
(TP14) Tillman, Pfetsch, IEEE Transactions on information processing, 2014.
(T15) Tillman, IEEE Signal Processing Letters, 2015.
(HS15) Hillar, Sommer, IEEE Transactions on information processing, 2015.

@@

As per the reviewer’s request, we have included in our updated manuscript a discussion of the computability of the sufficient conditions (they are, indeed, computable). The main constants in our work -- C_1, C_2 -- are computable, elementarily, in terms of two positive quantities that are natural in our context: (a) the restricted "Lower Bound" L_k(M) of a matrix M, which is a measure of how distinguishable two (k/2)-sparse x’s are by their images under M, and (b) the value in Def. 4 of \xi(V_1,..,V_l) given subspaces V_i spanned by columns of A, which bounds the distance between intersections of such subspaces and each subspace individually. The first quantity (a) is, in some form or another, related to standard objects in the field and may be computed as the smallest singular value of a set of specific sub-matrices; in particular, for fixed sparsity constraint k, the lower bound L_k(A) is efficiently computable (polynomial time). Because deciding L_k(A) > 0 is already NP-hard (TP14), it seems likely that without fixing a sparsity level there is no efficient computation in general for (a). The calculation of (b) involves determining "canonical angles" between subspaces, which is usually reduced again to a numerical (efficient) singular value decomposition given the fixed subspaces. However, upon close inspection, C_2 nonetheless requires an exponential number of instances of computing \xi. Thus, while our constants are effective (an algorithm exists to compute them), they are not necessarily polynomial-time computable.

We next respond to the reviewer's question about the mathematical "tightness" for the three types of quantities in our results : (a) required coding error eps(delta_1, delta_2) sufficient to guarantee uniqueness in terms of desired dictionary delta_1 and sparse representation error delta_2, (b) the constants C_1 and C_2 involved in our sufficient linear bound on eps(delta_1, delta_2) in terms of the noise, and (c) the number of sufficient samples N guaranteeing uniqueness.  With respect to (a), even with fixed dictionary, the recovery of sparse representations up to an error that is linear in the noise is essentially optimal (ACD13) (up to constants). Thus, the fact that we have a sufficient error in coding for uniqueness being linear for both the desired dictionary and sparse vector recovery error is optimal.  Although optimal, our solution to Problem 1 involves constants C_1 and C_2, and it would be interesting and relevant for practical applications to determine the minimal such constants. In the case of sparsity k=1, we have given an example showing optimality of constants in this case. Finally, with respect to (c), we regard this as a fundamental open problem in the field: how many sparsely generated samples determine the linear coding model?  (One reason to be skeptical that a polynomial suffice is again (TP14, T15).)  However, there is some good news -- if k is fixed or if the sparse codes \bar x_i are known to come from a polynomial number of supports, then a polynomial number of samples N is enough to guarantee uniqueness. We have revised the manuscript to discuss this last fact (which, incidentally, does not follow from the results of (HS15) in the noiseless case).

@@

Given a dictionary A and codes x_i that solve Problem 1 (e.g. learned by any algorithm), check that that they satisfy the assumptions of Thm. 1:
   - List the support sets of the x_i 
      - Discard those for which there are (k-1){m \choose k} or fewer x_i with that support.
      - Discard those for which the supported x_i are not in general linear position. (This should also exclude any supports less than k in size.)
   - From the support sets that remain, list all SIP-satisfying hypergraphs that are subsets of this set
   - Discard those for which L_{2H}(A) = 0.
   - For each of the remaining hypergraphs, determine the constant C_1 from A and the x_i (or for every subset of the x_i of size (k-1){m \choose k})
   - Check that the derived upper bound on \varepsilon exceeds that of the original problem. If yes, the solution is "unique".

There are other practical (polynomial time) implications of our discovery of sufficient combinatorial designs for support sets of generating codes x_i. Moreover, we have shown that a subset of dictionary elements are recoverable even if the number of dictionary elements in total is unknown; these observations are discussed in more detail below.

The reviewer also raises the concern that, regardless of practical implications, our results may amount to an incremental advance over those of (HS15) by way of similar techniques. We disagree with this assessment on both counts: our main result (Thm. 1) goes far beyond a straightforward extension of that in (HS15) to the noisy case and this required a significant deviation from their approach.

It is understandable that the reviewer may have thought otherwise, considering how the proof of Thm. 1 is presented in the manuscript. As observed by our second reviewer, however, the extension to the noisy case did indeed require a novel combination of several results in the literature. Specifically, the main difficulty was to generalize Lemma 1 to the case where the k-dimensional subspaces spanned by corresponding (through the map \pi) submatrices of A and B are assumed only to be proximal (small ‘d’), and not identical as in (HS15). In contrast to the noiseless case, here it must be explicitly demonstrated that this proximity relation is propagated through the repeated intersections of these submatrix spans all the way down to the spans of dictionary elements themselves. We designed and proved Lemma 4 to address this issue, which draws its bound from the convergence guarantees of an alternating projections algorithm first proposed by von Neumann. This result, combined with a little known fact about the distance ‘d’ requiring proof in (M10), constitute the more obscure components of the deduction in Eq. (26). To reiterate, this step is completely trivial in the noiseless case and required no mention for the inductive steps taken in (HS15).

### 2017-08-21 manuscripts/PNAS/chazthm_pnas.tex
added linear optimality ref

We close this section with a comment on sample complexity. A basic open problem remains: how many samples are necessary to determine the sparse linear coding model? If $k$ is held fixed or if the size of the support set of reconstructing codes is known to be polynomial in $\bar m$ and $k$, then a practical (polynomial) amount of data suffices.\footnote{In the latter case, a reexamination of the pigeonholing argument in the proof of Thm.~\ref{DeterministicUniquenessTheorem} requires a polynomial number of samples distributed over a polynomial number of supports.} Reasons to be skeptical that this holds in general, however, can be found in \cite{tillmann2014computational, Tillmann15}.

@@

In fact, uniqueness guarantees with minimal assumptions apply to all areas of data science and engineering that utilize learned sparse structure. For example, several groups have applied compressed sensing to signal processing tasks: MRI analysis \cite{lustig2008compressed}, image compression \cite{Duarte08}, and, more recently, the design of an ultrafast camera \cite{Gao14}. Given such effective uses of compressed sensing, it is only a matter of time before these systems incorporate dictionary learning to encode and process data (e.g., in a device that learns structure from motion \cite{kong2016prior}). In these cases, assurances such as those offered by our theorems certify that different devices (with different initialization, samples, etc.) will learn equivalent representations given enough data from statistically identical systems.\footnote{To contrast with the current hot topic of ``Deep Learning'', there are few such uniqueness guarantees for these multi-layer feedforward models of data; moreover, there is evidence that even minuscule amounts of (``adversarial'') signal noise can dramatically alter their output \cite{goodfellow2014explaining}.} Indeed, it seems a main reason for the sustained interest in dictionary learning as an unsupervised method for data analysis appears to be the assumed well-posedness of parameter identification in the sparse linear coding model, confirmation of which forms the core of our theoretical findings.

@@

In this section, we prove Lem.~\ref{MainLemma} after stating and proving some auxiliary lemmas. 
%First, note that since $\|\mathbf{A}\mathbf{x}\|_2 \leq \max_j\|\mathbf{A}_j\|_2\|\mathbf{x}\|_1$ and $\|\mathbf{x}\|_1 \leq \sqrt{k} \|\mathbf{x}\|_2$ for $k$-sparse $\mathbf{x}$, by \eqref{Ldef} we have the following frequently applied inequality:
%\begin{align}\label{delrho}
%L_{k'}(\mathbf{A}) \leq  \max_j\|\mathbf{A}_j\|_2 \ \ \text{for all $k' \leq 2k$}.
%\end{align}

>>>>

We close this section with comments on optimality.  Our linear scaling for $\varepsilon$ in \eqref{epsdel} is essentially optimal (e.g., see \cite{arias2013fundamental}), but a basic open problem remains: how many samples are necessary to determine the sparse linear coding model? If $k$ is held fixed or if the size of the support set of reconstructing codes is known to be polynomial in $\bar m$ and $k$, then a practical (polynomial) amount of data suffices.\footnote{In the latter case, a reexamination of the pigeonholing argument in the proof of Thm.~\ref{DeterministicUniquenessTheorem} requires a polynomial number of samples distributed over a polynomial number of supports.} Reasons to be skeptical that this holds in general, however, can be found in \cite{tillmann2014computational, Tillmann15}.

@@

In fact, uniqueness guarantees with minimal assumptions apply to all areas of data science and engineering that utilize learned sparse structure. For example, several groups have applied compressed sensing to signal processing tasks: MRI analysis \cite{lustig2008compressed}, image compression \cite{Duarte08}, and, more recently, the design of an ultrafast camera \cite{Gao14}. Given such effective uses of compressed sensing, it is only a matter of time before these systems incorporate dictionary learning to encode and process data (e.g., in a device that learns structure from motion \cite{kong2016prior}). Assurances such as those offered by our theorems certify that different devices (with different initialization, samples, etc.) will learn equivalent representations given enough data from statistically identical systems.\footnote{To contrast with the current hot topic of ``Deep Learning'', there are few such uniqueness guarantees for these multi-layer feedforward models of data; moreover, there is evidence that even minuscule amounts of (``adversarial'') signal noise can dramatically alter their output \cite{goodfellow2014explaining}.} Indeed, it seems a main reason for the sustained interest in dictionary learning as an unsupervised method for data analysis is the assumed well-posedness of parameter identification in the sparse linear coding model, confirmation of which forms the core of our theoretical findings.

@@

In this section, we prove Lem.~\ref{MainLemma} and some auxiliary lemmas.  % stating and proving some auxiliary lemmas. 
%First, note that since $\|\mathbf{A}\mathbf{x}\|_2 \leq \max_j\|\mathbf{A}_j\|_2\|\mathbf{x}\|_1$ and $\|\mathbf{x}\|_1 \leq \sqrt{k} \|\mathbf{x}\|_2$ for $k$-sparse $\mathbf{x}$, by \eqref{Ldef} we have the following frequently applied inequality:
%\begin{align}\label{delrho}
%L_{k'}(\mathbf{A}) \leq  \max_j\|\mathbf{A}_j\|_2 \ \ \text{for all $k' \leq 2k$}.
%\end{align}
